Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-17 talk/dls2012/paper.tex
address review comments by mike pall

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimizations
are simple, requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
Especially for dynamic languages that typically perform quite a lot of loop invariant
type checking, boxed value unwrapping and virtual method lookups.
In this paper we explain a scheme pioneered within the context of the LuaJIT project
for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves.
We have implemented the scheme in PyPy's tracing JIT compiler,
where it can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

Mike Pall pioneered a solution to address this problem in the context of a
dynamic language using a tracing JIT compiler. He published his algorithm and
its rationale in 2009~\cite{pall_luajit_2009} and implemented it in LuaJIT
2.0\footnote{\texttt{http://luajit.org/}}, an open source JIT compiler for the Lua
language. His approach allows to reuse all forward pass
optimizations to achieve loop invariant code motion and other loop-related
optimizations, which greatly simplifies the implementation. Using this scheme
one does not need to change the underlying optimization much to get these
advantages. We have implemented the same approach in PyPy's tracing JIT
compiler the results of which we present here.

@@

\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|r|}
\hline
 & CPython & PyPy  & PyPy & LuaJIT & LuaJIT & GCC \\
 &         & no LP &      & no LP  &        & -O3 \\
\hline
FFT(1024,32768) & 469.07 & 20.83 $\pm$ 0.039 & 12.73 $\pm$ 0.029 & 4.42 $\pm$ 0.017 & 2.76 $\pm$ 0.017 & 1.40 $\pm$ 0.082\\
\hline
FFT(1048576,2) & 58.93 & 4.12 $\pm$ 0.020 & 2.05 $\pm$ 0.007 & 1.33 $\pm$ 0.028 & 1.08 $\pm$ 0.014 & 0.83 $\pm$ 0.044\\
\hline
LU(100,4096) & 1974.14 & 32.22 $\pm$ 0.281 & 13.39 $\pm$ 0.063 & 8.57 $\pm$ 0.012 & 1.52 $\pm$ 0.014 & 1.33 $\pm$ 0.070\\
\hline
LU(1000,2) & 955.31 & 14.98 $\pm$ 0.436 & 5.99 $\pm$ 0.416 & 3.99 $\pm$ 0.014 & 0.67 $\pm$ 0.010 & 0.65 $\pm$ 0.077\\
\hline
MonteCarlo(268435456) & 618.89 & 20.60 $\pm$ 0.097 & 15.33 $\pm$ 0.163 & 3.92 $\pm$ 0.016 & 2.82 $\pm$ 0.009 & 1.69 $\pm$ 0.096\\
\hline
SOR(100,32768) & 1458.12 & 8.24 $\pm$ 0.002 & 2.66 $\pm$ 0.002 & 2.02 $\pm$ 0.009 & 1.31 $\pm$ 0.010 & 1.76 $\pm$ 0.088\\
\hline
SOR(1000,256) & 1210.45 & 6.48 $\pm$ 0.007 & 2.10 $\pm$ 0.005 & 1.63 $\pm$ 0.009 & 1.08 $\pm$ 0.010 & 1.49 $\pm$ 0.042\\
\hline
SparseMatMult(1000,5000,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.64 $\pm$ 0.032 & 4.53 $\pm$ 0.032 & 1.84 $\pm$ 0.061\\
\hline
SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.016 & 2.42 $\pm$ 0.010 & 1.20 $\pm$ 0.053\\
\hline
\hline
conv3(1e5) & 50.14 & 1.09 $\pm$ 0.022 & 0.49 $\pm$ 0.028 & 0.67 $\pm$ 0.010 & 0.12 $\pm$ 0.010 & 0.52 $\pm$ 0.084\\
\hline
conv3(1e6) & 49.20 & 1.13 $\pm$ 0.043 & 0.51 $\pm$ 0.008 & 0.70 $\pm$ 0.008 & 0.18 $\pm$ 0.000 & 0.60 $\pm$ 0.064\\
\hline
conv3x3(1000,1000) & 138.95 & 0.70 $\pm$ 0.007 & 0.20 $\pm$ 0.009 & 0.49 $\pm$ 0.010 & 0.09 $\pm$ 0.010 & 0.17 $\pm$ 0.079\\
\hline
conv3x3(1000000,3) & 139.81 & 0.70 $\pm$ 0.005 & 0.21 $\pm$ 0.006 & 0.53 $\pm$ 0.008 & 0.13 $\pm$ 0.009 & 0.19 $\pm$ 0.061\\
\hline
conv5(1e5) & 74.65 & 1.22 $\pm$ 0.009 & 0.64 $\pm$ 0.005 & 0.84 $\pm$ 0.006 & 0.17 $\pm$ 0.010 & 0.55 $\pm$ 0.047\\
\hline
conv5(1e6) & 77.94 & 1.26 $\pm$ 0.009 & 0.68 $\pm$ 0.014 & 0.87 $\pm$ 0.010 & 0.21 $\pm$ 0.013 & 0.58 $\pm$ 0.049\\
\hline
dilate3x3(1000,1000) & 137.52 & 4.35 $\pm$ 0.014 & 3.91 $\pm$ 0.037 & 0.48 $\pm$ 0.014 & 0.09 $\pm$ 0.006 & 0.17 $\pm$ 0.061\\
\hline
sobel(1000,1000) & 104.02 & 0.49 $\pm$ 0.009 & 0.21 $\pm$ 0.004 & 0.60 $\pm$ 0.012 & 0.24 $\pm$ 0.018 & 0.17 $\pm$ 0.061\\
\hline
sqrt(float) & 14.99 & 1.37 $\pm$ 0.001 & 0.89 $\pm$ 0.000 & 1.06 $\pm$ 0.013 & 0.83 $\pm$ 0.010 & 0.85 $\pm$ 0.088\\
\hline
sqrt(int) & 13.91 & 3.22 $\pm$ 0.033 & 2.65 $\pm$ 0.001 & - & - & 1.25 $\pm$ 0.053\\
\hline
sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & - & - & 1.34 $\pm$ 0.061\\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark results in seconds with 95\% confidence intervals. Arrays of length $10^5$ and
  $10^6$ and matrices of size $1000\times 1000$ and $1000000 \times
  3$ are used. The one used in each benchmark is indicated in
  the leftmost column. For the matrices, only the number of rows are
  specified.} 
\end{figure*}

@@

The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of RPython's JIT machine code backend as well as missing
optimizations, like instruction scheduling. Also, in case of nested loops, 
operations are only moved out of the 
innermost loop. That is an issue when the innermost loop is 
short and a significant amount of time is spent in the outer loops. This is the case 
with for example SparseMatMult.

The speedups that LuaJIT gains from the loop optimization pass are similar to
those PyPy gains. In general, LuaJIT is even closer to C performance, sometimes
even surpassing it. LuaJIT is generating machine code of higher quality because
it has a much better register allocator than PyPy, among other things.

@@

\acks
We would like to thank Samuele Pedroni, Sven Hager and the anonymous reviewers
for helpful comments on drafts of this paper. We owe deep gratitude to Mike Pall
for making his impressive work on LuaJIT available and for detailed help on a
draft of the paper.

>>>>

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimizations
are simple, requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
Especially for dynamic languages that typically perform quite a lot of loop invariant
type checking, boxed value unwrapping and virtual method lookups.

In this paper we explain a scheme pioneered within the context of the LuaJIT project
for making basic optimizations loop-aware by
using a simple pre-processing step on the trace without changing the
optimizations themselves.

We have implemented the scheme in PyPy's tracing JIT compiler,
where it can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance into the ballpark of static language compilers.
\end{abstract}

@@

Mike Pall pioneered a solution to address this problem in the context of a
dynamic language using a tracing JIT compiler. He published his algorithm and
its rationale in 2009~\cite{pall_luajit_2009} and implemented it in LuaJIT
2.0\footnote{\texttt{http://luajit.org/}}, an open source JIT compiler for the Lua
language. His approach allows to reuse all forward pass
optimizations to achieve loop invariant code motion and other loop-related
optimizations, which greatly simplifies the implementation. We have implemented
the same approach in PyPy's tracing JIT compiler, the results of which we
present here.

@@

\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|r|}
\hline
 & CPython & PyPy  & PyPy & LuaJIT & LuaJIT & GCC \\
 &         & no LP &      & no LP  &        & -O3 \\
\hline
FFT(1024,32768) & 469.07 & 20.83 $\pm$ 0.039 & 12.73 $\pm$ 0.029 & 4.42 $\pm$ 0.017 & 2.76 $\pm$ 0.017 & 1.40 $\pm$ 0.082\\
\hline
FFT(1048576,2) & 58.93 & 4.12 $\pm$ 0.020 & 2.05 $\pm$ 0.007 & 1.33 $\pm$ 0.028 & 1.08 $\pm$ 0.014 & 0.83 $\pm$ 0.044\\
\hline
LU(100,4096) & 1974.14 & 32.22 $\pm$ 0.281 & 13.39 $\pm$ 0.063 & 8.57 $\pm$ 0.012 & 1.52 $\pm$ 0.014 & 1.33 $\pm$ 0.070\\
\hline
LU(1000,2) & 955.31 & 14.98 $\pm$ 0.436 & 5.99 $\pm$ 0.416 & 3.99 $\pm$ 0.014 & 0.67 $\pm$ 0.010 & 0.65 $\pm$ 0.077\\
\hline
MonteCarlo(268435456) & 618.89 & 20.60 $\pm$ 0.097 & 15.33 $\pm$ 0.163 & 3.92 $\pm$ 0.016 & 2.82 $\pm$ 0.009 & 1.69 $\pm$ 0.096\\
\hline
SOR(100,32768) & 1458.12 & 8.24 $\pm$ 0.002 & 2.66 $\pm$ 0.002 & 2.02 $\pm$ 0.009 & 1.31 $\pm$ 0.010 & 1.76 $\pm$ 0.088\\
\hline
SOR(1000,256) & 1210.45 & 6.48 $\pm$ 0.007 & 2.10 $\pm$ 0.005 & 1.63 $\pm$ 0.009 & 1.08 $\pm$ 0.010 & 1.49 $\pm$ 0.042\\
\hline
SparseMatMult(1000,5000,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.64 $\pm$ 0.032 & 4.53 $\pm$ 0.032 & 1.84 $\pm$ 0.061\\
\hline
SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.016 & 2.42 $\pm$ 0.010 & 1.20 $\pm$ 0.053\\
\hline
\hline
conv3(1e5) & 50.14 & 1.09 $\pm$ 0.022 & 0.49 $\pm$ 0.028 & 0.67 $\pm$ 0.010 & 0.12 $\pm$ 0.010 & 0.52 $\pm$ 0.084\\
\hline
conv3(1e6) & 49.20 & 1.13 $\pm$ 0.043 & 0.51 $\pm$ 0.008 & 0.70 $\pm$ 0.008 & 0.18 $\pm$ 0.000 & 0.60 $\pm$ 0.064\\
\hline
conv3x3(1000,1000) & 138.95 & 0.70 $\pm$ 0.007 & 0.20 $\pm$ 0.009 & 0.49 $\pm$ 0.010 & 0.09 $\pm$ 0.010 & 0.17 $\pm$ 0.079\\
\hline
conv3x3(1000000,3) & 139.81 & 0.70 $\pm$ 0.005 & 0.21 $\pm$ 0.006 & 0.53 $\pm$ 0.008 & 0.13 $\pm$ 0.009 & 0.19 $\pm$ 0.061\\
\hline
conv5(1e5) & 74.65 & 1.22 $\pm$ 0.009 & 0.64 $\pm$ 0.005 & 0.84 $\pm$ 0.006 & 0.17 $\pm$ 0.010 & 0.55 $\pm$ 0.047\\
\hline
conv5(1e6) & 77.94 & 1.26 $\pm$ 0.009 & 0.68 $\pm$ 0.014 & 0.87 $\pm$ 0.010 & 0.21 $\pm$ 0.013 & 0.58 $\pm$ 0.049\\
\hline
dilate3x3(1000,1000) & 137.52 & 4.35 $\pm$ 0.014 & 3.91 $\pm$ 0.037 & 0.48 $\pm$ 0.014 & 0.09 $\pm$ 0.006 & 0.17 $\pm$ 0.061\\
\hline
sobel(1000,1000) & 104.02 & 0.49 $\pm$ 0.009 & 0.21 $\pm$ 0.004 & 0.60 $\pm$ 0.012 & 0.24 $\pm$ 0.018 & 0.17 $\pm$ 0.061\\
\hline
sqrt(float) & 14.99 & 1.37 $\pm$ 0.001 & 0.89 $\pm$ 0.000 & 1.06 $\pm$ 0.013 & 0.83 $\pm$ 0.010 & 0.85 $\pm$ 0.088\\
\hline
sqrt(int) & 13.91 & 3.22 $\pm$ 0.033 & 2.65 $\pm$ 0.001 & - & - & 1.25 $\pm$ 0.053\\
\hline
sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & - & - & 1.34 $\pm$ 0.061\\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark results in seconds with 95\% confidence intervals. The leftmost column gives the
name of each benchmark and the values of the benchmark parameters used. The different benchmarks and the meaning of their parameters are described in Section~\ref{sec:benchmarks}.}
\end{figure*}

@@

The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of RPython's JIT machine code backend and the naive register allocator.
Also, in case of nested loops,
operations are only moved out of the 
innermost loop. That is an issue when the innermost loop is 
short and a significant amount of time is spent in the outer loops. This is the case 
with for example SparseMatMult.

The speedups that LuaJIT gains from the loop optimization pass are similar to
those PyPy gains. In general, LuaJIT is even closer to C performance, sometimes
even surpassing it. LuaJIT is generating machine code of higher quality because
it has more optimizations\footnote{See
\texttt{http://wiki.luajit.org/Optimizations}} and is a produces much better
machine code than PyPy.

@@

\acks
We would like to thank Samuele Pedroni, Sven Hager and the anonymous reviewers
for helpful comments on drafts of this paper. We owe gratitude to Mike Pall
for making his impressive work on LuaJIT publicly available and for detailed
reviews on drafts of the paper.
